Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
motivating MRfy

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\alignwidth{\ensuremath R} % number of columns in an alignment

@@

\subsection{Solving new problems using perspicuous Haskell}

@@

A hidden Markov model effectively uses local information about
amino-acid sequences.
But real proteins are three-dimensional structures,
and their amino-acid sequences are subject to \emph{nonlocal}
constraints. 
And when a real protein folds, groups of amino acids
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
We~are studying adjacent groups called \emph{beta strands}, which
can become hydrogen-bonded to each other,
and are effectively ``stuck together.''

During the course of evolution, a mutation in an amino acid that participates
in this beta-strand pairing may cause a protein to fail to fold into a useful
structure, unless that mutation is accompanied by a \textit{compensatory}
mutation in the amino acid with which it is paired. 
Thus, homologous
proteins---those that share structure and function---tend to \textit{conserve}
such pairwise-conditional amino acids.

In order to model these non-local interactions, computational biologists
have extended hidden Markov models to include these non-local interactions,
modeled as a \textit{dependency graph}, which transforms the model into a
\textit{Markov random field}. These approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

SMURF~\cite{Menke:2010ti} and SMURFLite~\cite{Daniels:2012}, our previous work,
compute the 
optimal parse of a protein sequence onto the directed graph representing a 
Markov random field using multidimensional dynamic programming.
Mathematically, we can represent the probability of a query sequence 
$x_{1}..x_{n}$ given a Markov random field with an additional term in the
Viterbi relations.
Specifically, each term $V_{j}^{\prime M}(i)$ becomes
$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log Pr(x_{i} | x_{\pi(j)})$, where
$x_{\pi(j)}$ is the amino acid in the query sequence that will be placed
in the beta-strand position paired with position $j$.
This nonlocal dependency prevents the problem from exhibiting optimal substructure,
preventing dynamic programming from solving the problem quickly.
SMURF's 
computational complexity quickly exploded with complex protein structures, 
while SMURFLite limited its computational complexity by also limiting the 
complexity of protein structure it would understand.
The algorithm underlying 
MRFy grew out of an understanding of these limitations.

Rather than attempt to compute exactly the optimal parse, MRFy treats the beta 
strands as ``beads'' which can be slid along a string.
In this analogy, the 
string is the sequence of amino acids of the query protein.
A given placement 
of these beads corresponds to an assignment of particular amino acid residues 
being hydrogen bonded to other particular amino acid residues, and thus MRFy 
computes a probability based on observed frequencies of residues hydrogen 
bonding in beta strands~\cite{Cowen:2002p588}.
In between these beta strands, the 
Markov random field is in fact a traditional hidden Markov model, whose optimal 
solution can be easily computed via the Viterbi algorithm.
MRFy performs a 
stochastic search to place these beta strands, based on an objective function 
that combines the probability computed by the Viterbi algorithm with that 
computed from the paired, hydrogen-bonded residues in the beta strands.
If a 
structure has many beta strands, then it also has many miniature hidden Markov 
models which can be solved via the Viterbi algorithm.

\begin{figure}[h!] 
\ifpdfmadness
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
\fi
\caption{The directed graph representing the states of a MRFy Markov random 
field finite state automaton.
Shaded nodes represent beta-strand positions which have been fixed into match states, 
while the dashed edges represent the long-range dependencies due to hydrogen 
bonding.
Note that in between the beta-strand regions are normal HMMs.}\label{mrf} \end{figure}

MRFy contains two distinct algorithms: the classical Viterbi 
dynamic-programming algorithm for hidden Markov models, and a stochastic search 
algorithm for placing beta-strand positions.
This stochastic search algorithm 
implements several search heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.

\subsection{Higher-order functions}

Just as the Viterbi algorithm is one major component of MRFy, stochastic search 
is the other.
In designing MRFy, we did not know in advance what sort of 
stochastic search technique would converge the fastest, or give us the best 
alignments with the fastest performance.
We decided to implement several 
different search strategies so that we could experiment with all of them.
We 
found this to be easy and pleasant thanks to Haskell's support of higher order 
functions.

@@

>>>>

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\alignwidth{\ensuremath R} % number of columns in an alignment
\newcommand\pairedwith[1]{{\pi(#1)}}

@@

\subsection{Exploring new algorithms using higher-order functions}

@@

A hidden Markov model uses local information about
amino-acid sequences.
But real proteins are three-dimensional,
and when a real protein folds, groups of amino acids
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
We~are studying adjacent groups called \emph{beta strands}, which
can become hydrogen-bonded to each other,
and are effectively ``stuck together.''

A~mutation in a paired beta strand may break the pairing, in which
 case the mutated protein will not fold properly and is unlikely to
 survive natural selection.
But a mutation in a paired beta strand \emph{can} survive if it is
 accompanied by a compensatory mutation in the strand with which it is
 paired. 
The probability of finding amino acid $x_i$ in column~$j$ of an alignment may
 therefore depend on the amino acid in the paired column
 $\pairedwith j$.
This nonlocal interaction changes
Viterbi's equations:
each term $V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
The distance between columns~$j$ and $\pairedwith j$ can be as small
 as a few columns or as large as a few hundreds of columns.

The approach currently taken in computational biology
is~to model each pair of beta strands not with two hidden Markov
models, 
but with a pair of sequences of match states.
Each of the subsequences \emph{between} paired beta strands is then
modeled with an ordinary hidden Markov model.
The~combined model, as shown in \figref{mrf}, is called a
\textit{Markov random field}. 
Because of nonlocal interactions between paired beta strands, 
the likelihood-maximization for a Markov random field does not
 exhibit optimal substructure, and dynamic
 programming cannot solve it quickly \cite{Menke:2010ti,Daniels:2012}.
The~purpose of MRFy is to explore new methods of \emph{approximating}
maximum likelihood.

\begin{figure}
\ifpdfmadness
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.pdf}} 
\else
\centerline{\includegraphics[width=8cm]{mrf_interleave_diagram.eps}} 
\fi
In this Markov random field,
each shaded node represents a beta-strand position.
Nodes connected by dashed edges are hydrogen-bonded.

\caption{A Markov random field with two beta-stand pairs}
\label{mrf} 
\end{figure}

MRFy treats the beta 
strands, which are build into the model,
 as ``beads'' which can be slid along the query sequence to determine
 a \emph{placement} of the beta strands.
A~placement's probability is computed
based on observed frequencies of paired amino acids in 
hydrogen-bonded beta strands~\cite{Cowen:2002p588}.
Given a placement, the maximum likelihood of any subsequence that appears
\emph{between} two beta strands can be computed exactly and
efficiently using Viterbi's algorithm.
(The result of the exact computation is a probability that is
\emph{conditioned} on the placement.)
MRFy searches for placements stochastically,
trying to maximize the product of these probabilities.
We~are exploring several
heuristics: random-mutation hill climbing, simulated 
annealing, multistart simulated annealing, and a genetic algorithm.
Our~exploration of these heuristics relies on higher-order functions.

@@

\vfill

\begingroup
\advance\leftskip by 0pt plus 10em
\emph{The Haskell wave passes, and if you slide off the crest, you
drown.}
\endgroup
